---
title: "Intro to R II"
subtitle: "Complexities in analyzing conflicts: Data wrangling and data management in R"
date: "November 8, 2019"
author: "Cosima Meyer"
output: pdf_document
---

## Recap: What did we learn last session?

---

- **R Universe**: R, R Studio, R Markdown

- Code needs to be written in chuncks (shortcut for chuncks: Alt + Cmd + I (Mac), Alt + Ctrl + I (Windows))

- R is **object-based**; that means we create and work with objects. Objects can be vectors, matrices, or data frames for example.

---

- **Vectors**:
  - One-dimensional array
  - Can take various data types: numeric, logical, character
  - All elements have the same data type

An example for vectors: 
```{r}
countries <- c("RWA", "AGO", "BEN") # This vector is consists of characters
countries # This command displays the vector
```

```{r}
years <- c(1990, 1990, 1991) # This vector is numeric
years # This command displays the vector
```

---

- **Matrices**:
  - Two-dimensional array
  - Can take various data types: numeric, logical, character
  - All elements have the same data type
  
An example for a matrix:
```{r}
mat_combined <- cbind(countries, years) # This command generates the matrix mat_combined
```

\pagebreak

## How to access values in a matrix?

To access a single value in a matrix, we need to think in a matrix. We need to call the matrix object first with `mat_combined` and then add with square brackets the rows, columns, or cells. The basic syntax is as follows `[row, column]`. If we just want to access the third row, we can write: 

```{r}
mat_combined[3,] # We access the third row
```

If we want to access the second column we need to write respectively:

```{r}
mat_combined[,2] # We access the second column
```

How do we then access the cell in the first column and first row?

```{r}

```

\pagebreak

## Data frames

Instead of creating matrices, we can also create a data frame. Data frames are a standard way of handling social science problems in R. We generate a data.frame using `data.frame()`.

```{r}
df_combined_test <- data.frame(countries, years)
```

We add `stringsAsFactors=FALSE` to avoid that R transforms our string variable `countries` in factors.

```{r}
df_combined <- data.frame(countries, years, stringsAsFactors = FALSE)
```

#### Hands-on exercise

Remember what you generated last session? I'll provide you with two vectors for `city` and `population`.

```{r}
# Vector for city
city <- c("Heidelberg", "Frankfurt am Main", "Hamburg", "Berlin", "Schwerin")

# Vector for population
population <- seq(10000, 100000, length.out = 5)
```

Repeat the same for your two vectors created above (`city` and `population`) and generate a data frame that does not transform strings into factors.

```{r}

```

\pagebreak 

### How to access parts of a data frame?

As in matrices, we can also access variables in data frames-- that are usually listed in the columns of a data frame -- with the dollar sign (`$`). 
Let's assume we want to access the column that contains information on countries in the data frame `df_combined`:

```{r}
df_combined$countries
```

Accessing single columns seems to be fairly simple. But what if you want to access the very first observation of a column? (This might be useful for later data wrangling exercises.) We will use functions from the [`dplyr`](https://dplyr.tidyverse.org) package here. 
```{r, message = FALSE}
# Load package
# install.packages("dplyr")
library(dplyr)
```

To access the first observation, we use `first()`.

```{r}
first(df_combined$countries)
```

If we want to access the last observation, we use `last()`.

```{r}
last(df_combined$countries)
```

Or, if we want to access the nth observation, we simply use `nth()`. Note, we need to specify which observation we want to display!

```{r}
nth(df_combined$countries, 2)
```

#### Bonus: What is the difference between a `data.frame` and a `data.table`?

A [data.table](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) is an advanced version (a different format of) a data.frame that allows you to work with larger data sets. The syntax is fairly similar to SQL and base R. I hardly ever work with data.tables -- I personally prefer the tidyverse because I find it more intuitive. We will learn more about this in our next session.

---

#### Summary 

- **Vectors**:
  - One-dimensional array
  - Can take various data types: numeric, logical, character
  - All elements have the same data type
- **Matrices**:
  - Two-dimensional array
  - Can take various data types: numeric, logical, character
  - All elements have the same data type
- **Data frames**:
  - Two-dimensional array
  - Can take various data types: numeric, logical, character
  - Within a column, all elements have the same data type; different columns can have different data types

\pagebreak

## How to structure a typical R document best

We will always come back to this structure, so please see it as a first disclaimer on how our documents will look like. 
I will also provide you with a template of this structure for your own research.

### 1. Preparation: Remove documents, install packages, load data sets
```{r, message=FALSE}
# Remove all objects from R's memory
rm(list=ls())  
```

```{r, message=FALSE, warnings=FALSE, results="hide"}
# We will use the following code to install all packages
# This code makes particularly sense if you require many packages.

packages <- c("ggplot2")

# Install uninstalled packages
lapply(packages[!(packages %in% installed.packages())], install.packages)

# Load all packages to library
lapply(packages, library, character.only = TRUE)
```

```{r, eval=FALSE}
# Set working directory
setwd() # This command sets the working directory 
# (not required if you are working in a project as we do right now)
getwd() # This command shows you the present working directory
dir() # This command shows you what's in your directory
```

### 2. Do the data management, analysis, visualization ...
```{r}
# Data management

# Data analysis

# Visualization

```

If you have more intensive data management and analysis projects, you might also consider saving the different sub-parts into single documents. For an illustrative example, see [here](https://bookdown.org/lyzhang10/lzhang_r_tips_book/how-to-organize-a-project-folder.html) and [here](https://nicercode.github.io/blog/2013-04-05-projects/).
